---
title: "Visualization of Missing Values and the Effect of Different Weather Variables"
author: "Anton Hung"
date: "2022-11-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading functions from a separate r script}
setwd('/Volumes/GoogleDrive/Mon disque/wrangling/project/wranglinghub')
source('functions_library/functions_library.R')
```

Loading the data
```{r}
library(tidyverse)
setwd('/Volumes/GoogleDrive/Mon disque/wrangling/project/wranglinghub')

data <- read_csv('football_data.csv')
head(data)
```

MISSING DATA
```{r}
summary(data)
```

Visualizing which columns have missing data. "weather_detail" has the most missing values by far. This variable contains 8 factors: "DOME", "DOME (Open Roof)", "Fog", "Rain", "Rain | Fog", "Snow", "Snow | Fog", and "Snow | Freezing Rain". All of these factors are related to poor weather, or dome-related changes. Values were not recorded for "nice" weather, which is probably contributing to the large amount of missing values. To proceed, we simply decided not to use this column in our analysis. There were many other weather-related columns available in the dataset. 

```{r}
data_missing <- data
library(naniar)
gg_miss_var(data_missing)

# levels(as.factor(data_missing$weather_detail))
```

```{r}
library(naniar)
vis_miss(data_missing) 
```

```{r}
library(naniar)

# mcar_test(data.frame(data)) # this runs into an error, but we can subset the data by columns and 
# run an mcar test on just a portion of the data

# this confirms that the data is missing not a random
mcar_test(data.frame(data[,10:20])) 
```

```{r}
library(naniar)

gg_miss_fct(x = data_missing, fct = schedule_season) 
# a lot of missing data related to betting up until near 1978
# some missing data in the 2023 season, where the authors of the dataset pre-filled data for some games that haven't been played yet.


```
Evidently, there is a pattern in our missing data. We have 100% missing data in the earlier football seasons. This makes sense because when football first emerged, betting was not yet available

```{r}
pct_miss(data)
betting_columns <- c("team_favorite_id",
                     "spread_favorite",
                     "over_under_line",
                     "Difference_favored_minus_notfavored",
                     "Actual difference - spread",
                     "Abs value of spread")
pct_miss(data[,betting_columns]) # 20.34%
pct_miss(data[, -which(colnames(data) %in% betting_columns)])
```

Looking at the data in excel, our regular season betting data begins at row 2494 (2493 if we do not include the header).

```{r}

pct_miss(data[1:2492,betting_columns])
pct_miss(data[2493:13248,betting_columns])
pct_miss(data[13249:nrow(data),betting_columns])

```

# WEATHER

```{r}
# weather_temperature
# weather_wind_mph
# weather_humidity

# stadium_weather_type
# stadium_surface

# Actual difference - spread

betting_data <- data[2493:13248,]


```

Playing surface:
```{r}
table(as.factor(betting_data$stadium_surface))

plot_playing_surface(betting_data)

```

```{r}
table(betting_data$stadium_weather_type)

plot_stadium_weather(betting_data)

```

Temperature
```{r}
# weather_temperature
# weather_wind_mph
# weather_humidity
domed_stadiums <- filter(betting_data, 
                       stadium_weather_type=='dome')

non_domed_stadiums <- filter(betting_data, 
                       stadium_weather_type=='cold' |
                         stadium_weather_type=='moderate' |
                         stadium_weather_type=='warm')

summary(non_domed_stadiums$weather_temperature) # 50,64,72
summary(non_domed_stadiums$weather_wind_mph) # 0, 7, 11
summary(non_domed_stadiums$weather_humidity) # 56, 68, 78
# new values (when removing domed stadiums):
# 45, 57, 67.5
# 6, 9, 12
# 57, 69, 78

ggplot(non_domed_stadiums, aes(x=weather_temperature, y=`Actual difference - spread`)) +
  geom_point()

ggplot(non_domed_stadiums, aes(x=weather_wind_mph, y=`Actual difference - spread`)) +
  geom_point()
  
ggplot(non_domed_stadiums, aes(x=weather_humidity, y=`Actual difference - spread`)) +
  geom_point()
```

Categorical weather:
```{r}
plot_weather_status(betting_data)

```

